Dissecting the relationship between protein structure and sequence variation

### 2015-05-21 manuscript/voronoi/mnv.tex
MN updates. Some Figures regenrated

\documentclass[11pt]{article}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{graphicx}
\usepackage{footnote}
\usepackage{rotating}
\usepackage{slashbox}
\usepackage{afterpage}
\usepackage{float}
\citestyle{mbe}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%force setting figures at the top of the page:
    \makeatletter
    \setlength{\@fptop}{0pt}
    \makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  personal abbreviations and macros
%    the following package contains macros used in this document:
%\input econfmacros.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textwidth=6.9in  \textheight=9.6in
%%  Adjust these for your printer:
\leftmargin=-5in   \topmargin=-0.8in
\hoffset = -0.95in  \voffset = 0pt
\marginparwidth = 0pt
%% preprint number data:
%\newcommand\pubnumber{Article 14 in eConf C1304143}
\newcommand\pubnumber{}
\newcommand\pubdate{\today}

@@

    Figure \ref{fig:voronoi_ER} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER). It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties. In general, it is observed that the cell surface volume and area have the best prediction power compared to other cell characteristics, followed by the cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy. \\

@@

    Not shown here for brevity, almost identical results to the above are obtained if sequence entropy as defined by Eqn. \ref{eqn:shannon} were used in place of sequence evolutionary rates. The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range inherent in the definition of sequence entropy. \\

    Our conclusion is that the {\it edge effects} due to Voronoi tessellation appear to have $\lesssim0.01$ influence on the observed sequence-structure correlations in the dataset of $209$ proteins considered in this work. Similar conclusions are reached if the open cells were alternatively ranked by different criteria such as the fractional changes in cell area (vs. cell volume) upon changing the box size. The Voronoi cell characteristics, in particular cell volume and cell area can be safely used in predicting sequence variability without recourse to corrections for edge effects. An exception however is cell sphericity as defined in Eqn. \ref{eqn:sphericity}, which turns out to behave differently for open and closed cells. This is well illustrated in the adjacent averaging plots of Figure \ref{fig:zr4s_JC_voro} in which the behavior open and closed Cell characteristics, averaged over all sites in all proteins in our dataset, are plotted against the $normalized$ sequence evolutionary rates. For comparison, Figure \ref{fig:zr4s_JC_nonvoro} depicts the general behavior of the normalized site-specific evolutionary rates versus site-specific sequence entropy, $\Delta\Delta G$ rate, RSA, WCN, average Side-Chain B factor, Hydrogen bond strengths. \\

@@

    Throughout the previous sections, a variety of site-specific structural characteristics of proteins were calculated and discussed, using different sets of atomic coordinates representing individual sites in proteins. These include the weighted contact number, the Voronoi cell characteristics, and representative site-specific B factor. To make comprehensive comparison of all site-specific structural properties of proteins, we also calculate the Relative Solvent Accessibility (RSA) and \ddg rate \citep[e.g.,][]{echave_relationship_2014} of individual amino acids in their sites in proteins. \\

    Figure \ref{fig:best_predictorER} \& \ref{fig:best_predictorSE} compare the predictive power of each of site-specific structural quantities about sequence variability as measured by evolutionary rates ({\it r4sJC}) and sequence entropy. It is notable that once WCN is recalculated using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific sequence variability. The better performance of WCN compared to local packing density as measured from Voronoi cell volume and area may not be surprising, since WCN by its definition takes into account the potential long-range interactions among amino acids in different regions of protein. Indeed, the fractal dimension of proteins very much resembles that of lattice percolation models \cite{stauffer_introduction_1994} and similarly the random packing of hard spheres near percolation threshold \cite{lorenz_universality_1993, liang_are_2001}. To expand on this, define the average maximum extent of a protein as,

    \begin{equation}
        \label{eqn:max_extent}
        R_m = \frac{1}{2d}\sum_{i=1}^d \big( x_{i,max} - x_{i,min} \big),
    \end{equation}

    \noindent in which $d=3$ is the dimension of the Euclidean space $x_{i,max} - x_{i,min}$ is the maximum physical extent of the protein, as represented by the geometric center of the side-chain coordinates, in each of the three spatial dimensions. Alternatively, the radius of gyration of a protein of length $N$ can be defined as (similar to that of a finite size cluster: Egn. 45a, Sec. 3.2, in \cite{stauffer_introduction_1994}),

    \begin{equation}
        \label{eqn:rg}
        R_g = \sqrt{\sum_{i=1}^N \frac{|\vec r_i-\vec r_0|^2}{N}},
    \end{equation}

    \noindent where,

    \begin{equation}
        \label{eqn:com}
        \vec r_0 = \sum_{i=1}^N \frac{\vec r_i}{N},
    \end{equation}

    \noindent is the geometric center of the protein, and $r_i$ is the position of the geometric center of the side-chain of the $i^{th}$ amino acid in protein. This definition is such that the kinetic energy and the angular momentum of the protein about the $\vec r_0$ is equivalent to the kinetic energy and the angular momentum of all amino acids residing on a ring of radius $R_g$ centered at $\vec r_0$. Figure \ref{fig:maxextent} \& \ref{fig:rg} depict the behavior of the maximum extent ($R_m$) versus protein volume ($V$) and the radius of gyration ($R_g$) versus the protein length ($N$), respectively. \\

    The protein volumes and surface areas calculated using {\it 3V} software \cite{voss_3v:_2010}. A symmetric linear fit (i.e., Deming regression) to the plot of $\log R_m$ vs. $\log V$ and $\log R_g$ vs. $\log N$ results in regression slopes of $D\simeq2.47\pm0.06$ \& $D\simeq2.60\pm0.08$ respectively. The observed exponents are very similar to those of the scaling relation,

    \begin{equation}
        V \propto R^{D}, ~~~~~ (p\simeq p_c)
    \end {equation}

    \noindent with $D\simeq2.51\pm0.01$ in numerical simulations of hard-sphere packing \cite{lorenz_universality_1993} in three dimensions near percolation threshold ($p\simeq p_c$) or the exponents derived from lattice models $D\simeq2.54\pm0.05$ \cite{adler_series_1990} \& $D\simeq2.5$ \cite{stauffer_introduction_1994}. It is notable that far from percolation threshold (i.e., $p\rightarrow 0$) in three dimensions, $D=2$ \cite{parisi_critical_1981}, while above the threshold ($p>p_c$), $D=d=3$, in 3D space.

@@

\appendix{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}

@@

    \appendix{Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
    \label{app:best_crd}

@@

@@

    \begin{figure}
        \begin{center}
        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_nonvoro/zr4s_JC_nonvoro.png}
        \end{center}
        \caption{General behavior of site-specific structural characteristics versus site-specific evolutionary rates among all sites in all 209 proteins in dataset. The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites. The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
        \label{fig:zr4s_JC_nonvoro}
    \end{figure}

@@

    \begin{figure}
        \begin{center}
        \includegraphics[width=5.5in]{best_structural_predictors_of_SE.pdf}
        \end{center}
        \caption{A comparison of the prediction power of five structural variables about site-specific Sequence Entropy (SE). All structural quantities correlate positively with SE, with the exception of Weighted Contact Number (WCN) which correlates negatively. For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure. Note that the Spearman's $\rho$ is a rank correlation coefficient, meaning that the use of inverse WCN only changes the sign and not the magnitude of $\rho$. The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
        \label{fig:best_predictorSE}
    \end{figure}

@@

    \begin{table}[htbp]
    \caption{Best free parameter values of the Contact Number (CN) and the Weighted Contact Number with power-law kernel (WCN) that result in the strongest median Spearman's correlation ($\rho$) of WCN with four site-specific quantities (average side-chain B factor, evolutionary rates (r4sJC), sequence entropy, and \ddg rate) for the entire dataset of 209 proteins. The corresponding median correlation coefficients ($\rho$) are reported  inside parenthesis next to each parameter value in the table. \label{tab:best_params}}
    \bigskip
    \centerline{
    \begin{tabular}{|l|c|c|c|c|}
      \hline
      \multirow{3}{*}{Correlation} & \multicolumn{4}{|c|}{Best Performing Free Parameter of the Kernel} \\ \cline{2-5}
                        & Power-law & Exponential & Gaussian & Cutoff \\
                        & $\alpha$ & $\lambda~[\AA]$ & $\sigma~[\AA]$ & $r~[\AA]$ \\ \hline
      WCN--B factor     & -2.2 ($0.70$) & 3.8 ($0.70$) & 9.6 ($0.70$) & 13.0 ($0.69$) \\
      WCN--r4sJC        & -2.3 ($0.64$) & 3.4 ($0.64$) & 9.8 ($0.63$) & 14.8 ($0.62$) \\
      WCN--Seq. Entropy & -2.2 ($0.54$) & 3.4 ($0.53$) & 8.0 ($0.53$) & 15.2 ($0.52$) \\
      WCN--\ddg Rate    & -3.4 ($0.76$) & 2.0 ($0.78$) & 5.4 ($0.79$) & 9.8 ($0.75$) \\
      \hline
    \end{tabular}}
    \end{table}

>>>>

\documentclass[11pt]{article}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{graphicx}
\usepackage{footnote}
\usepackage{rotating}
\usepackage{slashbox}
\usepackage{afterpage}
\usepackage{float}
\usepackage{rotating}
\usepackage{url}
\usepackage{float}
\usepackage{multirow}
\usepackage{rotating}

\citestyle{mbe}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%force setting figures at the top of the page:
    \makeatletter
    \setlength{\@fptop}{0pt}
    \makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  personal abbreviations and macros
%    the following package contains macros used in this document:
%\input econfmacros.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textwidth=6.9in  \textheight=9.6in
%%  Adjust these for your printer:
\leftmargin=-5in   \topmargin=-0.8in
\hoffset = -0.95in  \voffset = 0pt
\marginparwidth = 0pt
%% preprint number data:
%\newcommand\pubnumber{Article 14 in eConf C1304143}
\newcommand\pubnumber{}
\newcommand\pubdate{\today}

@@

    Figure \ref{fig:voronoi_ER} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER). It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties. In general, it is observed that the cell volume and surface area have the best prediction power compared to other cell characteristics, followed by the cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy. \\

@@

    Not shown here for brevity, almost identical results to the above are obtained were sequence entropy as defined by Eqn. \ref{eqn:shannon} used in place of sequence evolutionary rates. The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range inherent in the definition of sequence entropy. A potential caveat with the Voronoi tessellation of finite structures such as proteins is the presence of {\it edge effects} in the properties of the cells that remain open, typically on the surface of the structure. We show in appendix \ref{app:edge_effects} that such effects are generally negligible in the results presented in this section. \\

@@

    Figure \ref{fig:best_predictorER} compares the predictive power of each of site-specific structural quantities about sequence variability as measured by evolutionary rates ({\it r4sJC}) for the dataset of $209$ monomeric enzymes. Not shown here for brevity, similar results are also obtained for sequence entropy used as the measure of sequence variability. For comparison, the results for WCN calculated using the C$_\alpha$ atomic coordinates are also illustrated in the plot, in addition to WCN calculated from the geometric center of the side chains as the representative coordinates of individual sites. Notably, the quantity WCN outperforms all other structural quantities -- including those derived from Voronoi tessellation -- in explaining site-specific sequence variability \footnote{The pairwise t-tests for all sets of correlations are available in the repository of the work (c.f., Section \ref{sec:methods})}. The better performance of WCN compared to local packing density as measured from Voronoi cell volume may not be surprising, knowing that WCN by its definition (with a power-law weighting function of exponent $\alpha=-2$) takes into account the potential long-range interactions among amino acids in different regions of protein. \\

    In order to segregate the combined effects of long-range interactions from local packing density in the relationship between WCN and sequence variability (e.g., as measured by r4sJC evolutionary rates), the inverse of Voronoi cell volume can be used as a maximally local measure of packing density. By controlling for the Voronoi cell volume, one can then quantify the residual influence of WCN -- that is, the the effects long-range amino acid interactions beyond the immediate neighbors -- on sequence variability. Figure

@@

\appendix

\section*{Appendix A: Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
\label{app:best_crd}

@@

    \section*{Appendix B: Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
    \label{app:best_crd}

@@

\section{Appendix C: Edge-Effects in Voronoi Partitioning of Protein Structures}
\label{app:edge_effects}

    Our conclusion is that the {\it edge effects} due to Voronoi tessellation appear to have $\lesssim0.01$ influence on the observed sequence-structure correlations in the dataset of $209$ proteins considered in this work. Similar conclusions are reached if the open cells were alternatively ranked by other criteria such as the fractional changes in cell area (vs. cell volume) upon changing the box size. The Voronoi cell characteristics, in particular cell volume and cell area can be safely used in predicting sequence variability without recourse to corrections for the edge effects. An exception however is cell sphericity as defined in Eqn. \ref{eqn:sphericity}, which turns out to behave differently for open and closed cells. This is well illustrated in the adjacent averaging plots of Figure \ref{fig:zr4s_JC_voro} in which the behavior open and closed Cell characteristics, averaged over all sites in all proteins in our dataset, are plotted against the $normalized$ sequence evolutionary rates. For comparison, Figure \ref{fig:zr4s_JC_nonvoro} depicts the general behavior of the normalized site-specific evolutionary rates versus site-specific sequence entropy, $\Delta\Delta G$ rate, RSA, WCN, average Side-Chain B factor, Hydrogen bond strengths.

@@

@@

    \begin{figure}
        \begin{center}
        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_nonvoro/zr4s_JC_nonvoro.png}
        \end{center}
        \caption{General behavior of site-specific structural characteristics versus normalized site-specific evolutionary rates among all sites in all 209 proteins in dataset. The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites. The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
        \label{fig:zr4s_JC_nonvoro}
    \end{figure}

%    \begin{figure}
%        \begin{center}
%        \includegraphics[width=5.5in]{best_structural_predictors_of_SE.pdf}
%        \end{center}
%          \caption{A comparison of the prediction power of five structural variables about site-specific Sequence Entropy (SE). All structural quantities correlate positively with SE, with the exception of Weighted Contact Number (WCN) which correlates negatively. For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure. Note that the Spearman's $\rho$ is a rank correlation coefficient, meaning that the use of inverse WCN only changes the sign and not the magnitude of $\rho$. The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
%        \label{fig:best_predictorSE}
%    \end{figure}

@@

    \begin{table}[htbp]
    \caption{Median best free parameter values of the Contact Number (CN) and the Weighted Contact Number with power-law kernel (WCN) that result in the strongest median Spearman's correlation ($\rho$) of CN \& WCN with three site-specific quantities (evolutionary rates (r4sJC), sequence entropy, and average side-chain B factor) in the entire dataset of 209 proteins. The $25\%$ \& $75\%$ quartile distances in the distribution of parameters from the median are given as superscripts and subscripts to the median values. \label{tab:best_params}}
    \bigskip
    \centerline{
    \begin{tabular}{|l|c|c|}
      \hline
      Correlation with  & $r_0[\AA]$ (CN)      & $\alpha$ (WCN)        \\ \hline
      r4sJC             & $14.3^{+5.3}_{-4.0}$ & $-2.3^{+0.8}_{-0.4}$  \\
      Seq. Entropy      & $12.4^{+5.5}_{-2.6}$ & $-2.2^{+0.8}_{-0.4}$  \\
      B factor          & $13.1^{+4.0}_{-2.2}$ & $-2.2^{+0.5}_{-0.4}$  \\
      \hline
    \end{tabular}}
    \end{table}
